---
title: "Report - Exploratory Data Analysis"
output: html_notebook
---

```{r echo=FALSE, results='hide'}

library(tidyverse)
library(ggthemes)
library(lubridate)

edx <- readRDS(file = "./rda/full_edx.rds")
validation <- readRDS (file = "./rda/full_Validation.rds")

#edx <- readRDS(file = "./rda/small_edx.rds")
#validation <- readRDS (file = "./rda/small_Validation.rds")

rating_levels <- c("0.5", "1", "1.5", "2", "2.5", "3", "3.5", "4", "4.5", "5")

```

***
# Explore variables
```{r}

summary(edx)
head(edx)

```

## Response Variable: Rating
```{r}

summary(edx$rating)

```

```{r}

edx %>% ggplot(aes(x=rating)) +
  geom_histogram(binwidth = .5, col="grey") +
  scale_x_continuous(breaks = seq(0, 5, .5)) +
  labs(title = "Notice fewer 1/2 Stars, WHY????") +
  theme_economist_white()



```
## Other interesting variables
```{r}

#Find number of unique movies rated, number of unique users, and number of movies in each genre.

mjb1 <- length(unique(edx$movieId))
mjb2 <- length(unique(edx$userId))

rated_movies_grouped_by_genre <- edx %>% 
  separate_rows(genres, sep = "\\|") %>% 
  group_by(genres) 

mjb4 <- nrow(rated_movies_grouped_by_genre)

```
In the dataset there are: 

- `r mjb2` unique users
- `r mjb1` unique movies rated
- `r mjb4` unique movie genres

Number of reviewed movies in each genre.  Note that a movie may be tagged with several genres, e.g., the movie "Boomerang" is classified as both a "Comedy" and a "Romance."
```{r}

edx %>% 
  separate_rows(genres, sep = "\\|") %>% 
  group_by(genres) %>% 
  summarise(count = n()) %>% 
  ggplot() +
    geom_col(aes(x = reorder(genres, -count), y = count)) +
    coord_flip() +
    xlab("Genre") +
    ylab("Number of Movies") +
    theme_economist_white() +
    theme(axis.text.y.left = element_text(hjust = 1),
          panel.grid.major = element_blank(),
          axis.ticks.y.left = element_line(),
          axis.ticks.x.bottom = element_blank())

```

Which movies did not have a genre listed? 
```{r}

edx %>% filter(genres == "(no genres listed)")

```


What are the top 10 most reviewed movies?
```{r}

edx %>% 
  group_by(title) %>% 
  summarise(count = n()) %>% 
  arrange(-count) %>% 
  top_n(10)

```
Who are the top 10 active reviewers?
```{r}

edx %>% 
  group_by(userId) %>% 
  summarise(count = n()) %>% 
  arrange(-count) %>% 
  top_n(10)

```
***
# Relationships (part 1)
## Numeric Correlations with Rating
No continuous numeric predictors to correlate with Rating; userId and movieId are discrete while timestamp is a date.

*** 
# Missing data, label encoding, and factorizing variables
```{r}
#
# make a copy of the data to modify so I don't have to rebuild if something goes bad.
#
edx_working <- edx

#
# First of all, I would like to see which variables contain missing values.
#
NAcol <- which(colSums(is.na(edx)) > 0)

```

There are `r length(which(colSums(is.na(edx)) > 0))` columns with at least one NA.

```{r}

# # Label encoding
# # There are no labels that are ordinal and should be encoded as a number.
# 
# #
# # Separate and factorize the genres
# #
# edx_working <- edx_working %>% 
#   separate_rows(genres, sep = "\\|") %>% 
#   mutate(genre = factor(genres))
# 
# #
# # Create a month-year from the timestamp
# #
# edx_working <- edx_working %>% 
#   mutate(month = round_date(as_datetime(timestamp), unit = "month"),
#          bimonth = round_date(as_datetime(timestamp), unit = "bimonth"),
#          qtr = round_date(as_datetime(timestamp), unit = "quarter"),
#          season = round_date(as_datetime(timestamp), unit = "season"),
#          year = round_date(as_datetime(timestamp), unit = "year"))
# 
# #
# # Factorizing the rating
# #
# # edx2 <- edx2 %>% 
# #   mutate(rating = factor(rating))
# 
# edx_working %>% filter(genres == "(no genres listed)")
# # "Pull My Daisy (1958)" by Kerouac
# # IMDB lists the genre as "Short"


```

***
# Relationships (part 2)
Movie Genre and rating - is there a genre bias?
```{r eval=FALSE}

library(tidyverse)

edx <- readRDS(file = "./rda/full_edx.rds")
edx_working <- edx

# combined genre and rating
edx_working %>% 
  group_by(genres) %>% 
  summarise(avg_rating = mean(rating)) %>% 
  ggplot() +
    geom_point(aes(x = reorder(genres, avg_rating), y = avg_rating)) +
    coord_flip()

# individual genre and rating
edx_working %>% 
  separate_rows(genres, sep = "\\|") %>% 
  group_by(genres) %>% 
  summarise(avg_rating = mean(rating)) %>% 
  ggplot() +
    geom_point(aes(x = reorder(genres, avg_rating), y = avg_rating)) +
    coord_flip() +
    ylim(c(0, 5))

```
Yes, there is a small genre-bias overall.

ANOVA
```{r eval=FALSE}
library(tidyverse)

rating_levels <- c("0.5", "1", "1.5", "2", "2.5", "3", "3.5", "4", "4.5", "5")

edx <- readRDS(file = "./rda/small_edx.rds")
edx_working <- edx %>% 
  mutate(rating = factor(rating, rating_levels))

summary(aov(rating ~ userId + movieId, data = edx_working))

```



How are users, genres, and ratings related?  Do different users have a different genre-bias?

```{r}

library(tidyverse)

edx <- readRDS(file = "./rda/small_edx.rds")
edx_working <- edx

# combined genre and rating
edx_working %>% 
  group_by(genres) %>% 
  summarise(avg_rating = mean(rating)) %>% 
  ggplot() +
    geom_point(aes(x = reorder(genres, avg_rating), y = avg_rating)) +
    coord_flip()

```

``` {r}


#
# Separate and factorize the genres
#
edx_working <- edx_working %>% 
  separate_rows(genres, sep = "\\|") %>% 
  mutate(genre = factor(genres))

edx_working %>% 
  ggplot() +
    geom_density(aes(x = rating, col = genre))



# edx_working %>% 
#   ggplot() +
#     geom_point(aes(y der(genres, -avg_rating), x = avg_rating)) +
#     xlim(0, 5)

# do users have extreme views of each genre
user_genre_affinity <- edx2 %>% 
  group_by(userId, genres) %>% 
  summarise(avg_rating = mean(rating)) %>% 
  filter(avg_rating < 1 | avg_rating > 4.75)

user_genre_affinity %>% ggplot() +
    geom_raster(aes(y = genres, x = userId, fill = avg_rating)) +
    scale_fill_distiller(palette = "Spectral")


```

Average rating of each user.  Red represents a user who rated all movies the same (sd == 0).
```{r}
# user and rating
mjb1 <- edx %>% 
  group_by(userId) %>% 
  summarise(avg_rating = mean(rating), std_dev = sd(rating))%>% 
  arrange(-avg_rating)

outliers <- mjb1 %>% 
  filter(std_dev == 0)

ggplot() +
  geom_point(data = mjb1, aes(x = userId, y = avg_rating), color = "grey") +
  geom_point(data = outliers, aes(x = userId, y = avg_rating), color = "red")  

```
Let's look at the ratings of only these users
```{r}

edx %>%
  filter(userId %in% outliers$userId) %>% 
  ggplot() +
    geom_point(aes(x = as.factor(reorder(userId, rating)), y = rating))

```

How do ratings change over time?
```{r}

#
# make a copy of the data to modify so I don't have to rebuild if something goes bad.
#
edx2 <- edx %>% 
  mutate(timestamp = as_datetime(timestamp))

edx2 %>% 
  filter(timestamp > ymd("2002-01-01") & 
         timestamp < ymd("2004-01-01")) %>% 
  ggplot(aes(x = timestamp, y = rating)) +
    geom_point() +
    labs(title = "1/2 star ratings start on 2003-05-15")

edx2 %>% 
  mutate(time_bin = round_date(timestamp, unit = "month")) %>% 
  group_by(time_bin) %>% 
  summarise(avg_rating = mean(rating)) %>% 
  ggplot(aes(x = time_bin, y = avg_rating)) +
    geom_point() +
    geom_smooth(se = FALSE) +
    labs(title = "Monthly average movie rating changes over time.") +
    theme_economist_white()  

```
